Efficient ancestry and mutation simulation with msprime 1.0

### 2021-08-11 paper.tex
aaron's affiliation
Minor edits in recombination section

% Franz
\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
% Gertjan, Konrad
\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
% Daniel
\affil[3]{Khoury College of Computer Sciences, Northeastern University}
% Graham
\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
% Simon, Dom, Ivan
\affil[5]{FIXME AARON's AFFILIATION}
% Georgia
\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
    University of Melbourne}
% Jerome, Yan, Wilder, Joe
\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
    University of Oxford}

@@

What about simulations with changing population size?
To understand how run time depends on demography
it helps to consider why run time is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid and inconsequential events that
lead to the quadratic run time.
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of run time we can estimate when this occurs.
Suppose that the maximum number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
At the maximum, coalescence and recombination rates are equal,
so a typical segment of ancestry will spend roughly half its time
in a lineage with at least one other such segment---and,
since such lineages carry at least two segments,
at most one-third of the lineages carry long trapped segments of ancestry.
Since the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
this implies that $\ell \approx L / 6$.
Setting the rates of recombination and common ancestor events
to be equal and solving for $M$,
we find that $M$ is roughly equal to $L N_e$.
The number of lineages then decreases gradually from this maximum
on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
Since the total rate of events when the maximum number of lineages is
present is roughly $L^2 N_e / 6$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.

What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively
evenly on the coalescent time scale.
Suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then run time will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the run time will depend critically on $T$---in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.

>>>>

% Franz
\affil[1]{Mathematical and Computational Population Genetics, University of T\"ubingen}
% Gertjan, Konrad
\affil[2]{Institute of Evolutionary Biology, The University of Edinburgh}
% Daniel
\affil[3]{Khoury College of Computer Sciences, Northeastern University}
% Graham
\affil[4]{Lundbeck GeoGenetics Centre, Globe Institute, University of Copenhagen}
% Aaron
\affil[5]{Department of Integrative Biology, University of Wisconsin--Madison}
% Georgia
\affil[6]{Melbourne Integrative Genomics, School of Mathematics and Statistics,
    University of Melbourne}
% Jerome, Yan, Wilder, Joe
\affil[7]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery,
    University of Oxford}

@@

What about simulations with changing population size?
To understand how run time depends on demography
it helps to consider why run time is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid and inconsequential events that
lead to the quadratic run time.
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out.
To get an idea of run time we can estimate when this balance occurs.
Suppose that the maximum number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
At the maximum, coalescence and recombination rates are equal,
so a typical segment of ancestry will spend roughly half its time
in a lineage with at least one other such segment---and,
since such lineages carry at least two segments,
at most one-third of the lineages carry long trapped segments of ancestry.
Since the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
this implies that $\ell \approx L / 6$.
Setting the rates of recombination and common ancestor events
to be equal and solving for $M$,
we find that $M$ is roughly equal to $L N_e$.
The number of lineages then decreases gradually from this maximum
on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
Since the total rate of events when the maximum number of lineages is
present is roughly $L^2 N_e / 6$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.

What does this tell us about run time for simulating time-varying population sizes?
The argument above implies that the work is spread out relatively
evenly on the coalescent time scale.
Suppose that population size today is $N_1$,
while $T$ generations ago it was $N_2$.
Does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then run time will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the run time will depend critically on $T$---in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.
